The impact of forest management on the temperature sensitivity of organic matter turnover

### 2024-11-12 Manuscript_draft/Draft.Rmd
new draft.rmd

To test these hypotheses, we have recorded data on soil respiration and microclimate every month in three replicated Quercus forests across Europe subjected to different forest management strategies during three years. We have analysed the sensitivity of soil respiration to microclimate (i.e., soil temperature and soil moisture) by constructing a mechanistic parametric model and compared to a non-parametric model (i.e., random forest).

# Material and methods

@@

@@

## Characterization of study sites
Tree density in France was very high compared to Spain and especially to Romania where the tree density was very low (Figure x). In contrast, the DBH was the lowest in France and the highest in Romania (Figure x). Spain showed intermediate values of both tree density and DBH (Figure x). For soil variables, in the untouched forest (control), the bulk density was the highest in Spain and the lowest in France. Spain had the highest pH while France and Romania had similar ranges of pH. The highest significant levels of C and N were found in Spain while the lowest C content was found in Romania. France in turn presented doubled C:N rates than the other two sites (Figure x). Total microbial biomass, fungal biomass and bacterial biomass peaked in France, followed by Spain and it was the lowest in Romania. Similarly, in general, activities of most enzymes were higher in France compared with Spain and Romania (Figure x).

##Effects of management disturbance treatments on soil physical-chemical and biological properties
The disturbance treatments barely affected the soil properties in Romania (Figure x). In France, the bulk density increased as the intensity of the disturbance did (i.e., from thinning with slash to clear-cut no slash) and in Spain it significantly increased in all the disturbance treatments compared with control (Figure x). The pH only significantly increased in both clear-cut treatments in Spain. Moreover, in Spain the N content decreased in the thinning and clear-cut without slash (Figure x). In France, compared with controls, C content increased in all the disturbance treatment, but in the clear-cut without slash that decreased (Figure x). In Spain, C content only decreased in the thinning-without slash treatment (Figure x). 
Similar to soil properties, microbial biomass, fungal biomass and bacterial biomass did not change with the disturbance treatment in Romania. In Spain, these variables significantly decreased in the disturbance treatments where the slash was removed and in France only in the clear-cut treatment where the slash was removed (Figure x). The disturbance treatments barely affected enzymatic activities in Romania and Spain, but generally decreased in clear-cut where slash was removed in France (Figure x).

![Boxplots of soil microbial biomass and enzymatic activities per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/enzymes_boxplots.png){width=80%}
![Boxplots of stand characteristics and soil physical-chemical properties per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/edaphic_boxplots.png){width=80%}

@@

![Climatic parameters (soil temperature and soil moisture) and respiration measured at the study sites (mean values per month and treatment after loess interpolation with span = 0.5). Please notice thot not every time series covers the same period of time. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/climatedata_means.png){width=60%}

As expected, in Spain the temperature variation was the lowest compared to the other two sites. In general, control plots showed the lowest thermal amplitude in all sites (i.e., with higher temperatures in winter and lower in summer). In Spain, the soil temperature increased as the intensity of the tree removal, mainly in summer, but the addition of slash buffered these increases. In France, the main effect on soil temperature was the tree removal, especially in summer and this was independent on the addition of slash. Soil temperature was not affected by disturbance and post-disturbance treatments in Romania. The observed effects of the disturbance and post-disturbance treatments on soil temperature were decreasing with time.
Soil moisture fluctuated inversely with temperature. In Spain, soil moisture was higher in the clear-cut with and without slash addition, except in the driest periods (from April to August). Similarly, in Romania, clear-cut increased soil moisture independently of slash addition. Contrarily, in France, soil moisture tended to decrease in the clear-cut treatment with and without slash (Figure x). 
Both in Spain and France, removal of slash decreased respiration rates, independently of tree cutting intensity, and thinning coupled with slash addition kept the respiration rates as the same level as controls. In Romania, effects of the disturbances were lower, however clear-cut increased respiration rates, especially when slash was added to the soil. At the end of the monitoring period, three years for Spain and two years for France and Romania after disturbance, respiration rates increased in the clear-cut with slash addition in the three study sites (Figure x).

## Benchmark models and parametric model fit
![Fit of the two benchmark machine learning models, linear (a) and random forest (b)](../Figures/fit_ML_validation.png){width=80%}

![Fit of the parametric model,  constant `E_a`](../Figures/fit_model_validation.png){width=60%}

## Parametric model posteriors, constant `E_a` 
The activation energy (Ea) and temperature sensitivity (Q10) values of SOM mineralization were consistently higher in Romania, followed by France and Spain. In general, the activation energy and Q10 values decreased with slash addition due to higher SOM quality compared with treatments in which slash was removed. This was especially evident in Spain for both thinning and clear-cut and in the clear treatment in France. The activation energy and Q10 in the treatments where slash was added did not differ from controls (Figure x, xa). These results were consistent when the activation energy was analysed as a function of time since disturbance and the activation energy systematically decreased in all treatments and all countries with time since disturbance (Figure x). The photosynthetic activity did not modify the sensitivity of soil respiration to temperature (peak and amplitude factors from the sine-function not affected) (Figure xb, c). However, the amplitude of the sine-function tended to be higher in France (Figure xc). The parameters of soil moisture function (a and b) did not change the response of the model either (Figure xd, e).

![Posteriors of the parameters varying as a function of treatments, `E_a` constant over time](../Figures/posteriors_boxplots.png){width=60%}

## Parametric model posteriors,  time-variant `E_a` 
![Posteriors of the `E_a` parameter varying as a function of treatments and sampling year](../Figures/timeseries_Ea_timevariant.png){width=60%}

@@

## Role of temperature and moisture in conditioning decomposition
Temperature and moisture seem the most important components in explaining the variance of the soil respiration. The seasonality scaling, although apparently differentiating slightly between sites, presents always very low amplitude posteriors, weighting on average only a 5% of the total measured respiration. It was more important in France (between 4.5% and 12.2%) and much less in Spain and Romania (between 2.0% and 3.2%). This seasonality might represent many different processes, but the most likely are seasonal processes connected with the above-ground vegetation and their inputs to the soil, which in turn would control microbial communities.

## Activation energy of different organic material

## Role of edaphic, ecological and microbiological variables in conditioning decomposition
Any other variable than soil temperature and moisture did not seem to play a pivotal role in explaining the variance of the soil respiration.

@@

# Appendix
![Posteriors of the scaling parameter, varying as a function of single plots](../Figures/posteriors_boxplots_q10.png){width=60%}

![Posteriors of the scaling parameter, varying as a function of single plots](../Figures/Appendix/posteriors_boxplots_A.png){width=60%}

![Posteriors of the parameters of the temperature function, plotted as density distributions](../Figures/Appendix/posteriors_temp.png)

![Posteriors of the parameters of the moisture function, plotted as density distributions](../Figures/Appendix/posteriors_seasonality.png)

![Posteriors of the parameters of the seasonality function, plotted as density distributions](../Figures/Appendix/posteriors_moisture.png)

![Ratio (expressed as %) between the mean amplitude parameter and the mean measured respiration by treatment](../Figures/Amplitude_ratio.png)

![Temperature scaling function](../Figures/Appendix/Temp_function.png){width=50%}

![Moisture scaling function](../Figures/Appendix/Moist_function.png){width=50%}

![Seasonality scaling function](../Figures/Appendix/Season_function.png){width=50%}

>>>>

To test these hypotheses, we have recorded data on soil respiration and microclimate every month in three replicated Quercus forests across Europe subjected to different forest management strategies during three years. We have analysed the sensitivity of soil respiration to microclimate (i.e., soil temperature and soil moisture) by constructing a mechanistic parametric model and compared to a non-parametric model (i.e., random forest). We have also explored the main drivers of variation in Q10 across different forests and forest managements disturbances.
# Material and methods

@@

## Temperature sensitivity (Q10)

@@

## Characterization of study sites
Tree density in France was very high compared to Spain and especially to Romania where the tree density was very low (Figure 1). In contrast, the DBH was the lowest in France and the highest in Romania (Figure 1). Spain showed intermediate values of both tree density and DBH (Figure 1). For soil variables, in the untouched forest (control), the bulk density was the highest in Spain and the lowest in France. Spain had the highest pH while France and Romania had similar ranges of pH. The highest significant levels of C and N were found in Spain while the lowest C content was found in Romania. France in turn presented doubled C:N rates than the other two sites (Figure 1). Total microbial biomass, fungal biomass and bacterial biomass peaked in France, followed by Spain and it was the lowest in Romania. Similarly, in general, activities of most enzymes were higher in France compared with Spain and Romania (Figure 2).

![Figure1. Boxplots of stand characteristics and soil physical-chemical properties per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/edaphic_boxplots.png){width=80%}

## Effects of management disturbance treatments on soil physical-chemical and biological properties
The disturbance treatments barely affected the soil properties in Romania (Figure 1). In France, the bulk density increased as the intensity of the disturbance did (i.e., from thinning with slash to clear-cut no slash) and in Spain it significantly increased in all the disturbance treatments compared with control (Figure 1). The pH only significantly increased in both clear-cut treatments in Spain. Moreover, in Spain the N content decreased in the thinning and clear-cut without slash (Figure 1). In France, compared with controls, C content increased in all the disturbance treatment, but in the clear-cut without slash that decreased (Figure 1). In Spain, C content only decreased in the thinning-without slash treatment (Figure 1). 
Similar to soil properties, microbial biomass, fungal biomass and bacterial biomass did not change with the disturbance treatment in Romania. In Spain, these variables significantly decreased in the disturbance treatments where the slash was removed and in France only in the clear-cut treatment where the slash was removed (Figure 2). The disturbance treatments barely affected enzymatic activities in Romania and Spain, but generally decreased in clear-cut where slash was removed in France (Figure 2).

![Figure2. Boxplots of soil microbial biomass and enzymatic activities per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/enzymes_boxplots.png){width=80%}

@@

As expected, in Spain the temperature variation was the lowest compared to the other two sites (Figure 3a). In general, control plots showed the lowest thermal amplitude in all sites (i.e., with higher temperatures in winter and lower in summer). In Spain, the soil temperature increased as the intensity of the tree removal, mainly in summer, but the addition of slash buffered these increases. In France, the main effect on soil temperature was the tree removal, especially in summer and this was independent on the addition of slash. Soil temperature was not affected by disturbance and post-disturbance treatments in Romania. The observed effects of the disturbance and post-disturbance treatments on soil temperature were decreasing with time.
Soil moisture fluctuated inversely with temperature. In Spain, soil moisture was higher in the clear-cut with and without slash addition, except in the driest periods (from April to August). Similarly, in Romania, clear-cut increased soil moisture independently of slash addition. Contrarily, in France, soil moisture tended to decrease in the clear-cut treatment with and without slash (Figure 3b). 
Both in Spain and France, removal of slash decreased respiration rates, independently of tree cutting intensity, and thinning coupled with slash addition kept the respiration rates as the same level as controls. In Romania, effects of the disturbances were lower, however clear-cut increased respiration rates, especially when slash was added to the soil. At the end of the monitoring period, three years for Spain and two years for France and Romania after disturbance, respiration rates increased in the clear-cut with slash addition in the three study sites (Figure 3c).

![Figure 3. Climatic parameters (soil temperature and soil moisture) and respiration measured at the study sites (mean values per month and treatment after loess interpolation with span = 0.5). Please notice thot not every time series covers the same period of time. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/climatedata_means.png){width=60%}

## Benchmark models and parametric model fit
![Figure 4. Fit of the two benchmark machine learning models, linear (a) and random forest (b)](../Figures/fit_ML_validation.png){width=80%}

![Figure 5. Fit of the parametric model,  constant `E_a`](../Figures/fit_model_validation.png){width=60%}

## Temperature sensitivity (Q10) and parametric model posteriors 
The temperature sensitivity (Q10) values of SOM mineralization were consistently higher in Romania (2.9 in controls), followed by France (2.7) and Spain (2.5) (Figure S4). The activation energy (Ea) followed the same pattern as Q10 (Figure 6a, Figure S4). In general, the activation energy and Q10 values decreased with slash addition due to higher SOM quality compared with treatments in which slash was removed. This was especially evident in Spain for both thinning and clear-cut and in the clear treatment in France. The activation energy and Q10 in the treatments where slash was added did not differ from controls (Figure 6a, Figure S4). These results were consistent when the activation energy was analysed as a function of time since disturbance and the activation energy systematically decreased in all treatments and all countries with time since disturbance (Figure 7). The photosynthetic activity did not modify the sensitivity of soil respiration to temperature (peak and amplitude factors from the sine-function not affected) (Figure 6b, c). However, the amplitude of the sine-function tended to be higher in France (Figure 6c). The parameters of soil moisture function (a and b) did not change the response of the model either (Figure 6d, e).

![Figure 6. Posteriors of the parameters from the parametric model varying as a function of treatments: (a) activation energy(Ea) for temperature, (b) peak and (c) amplitude parameters from the seasonality sine-function, (d) and (e) correspond to a and b parameters from the soil moisture function, all testing the effects on respiration. Ea constant over time. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/posteriors_boxplots.png){width=60%}

![Figure 7. Posteriors of the activation energy (Ea) parameter varying as a function of treatments and sampling year per country](../Figures/timeseries_Ea_timevariant.png){width=60%}

The variables that most variance explained of Q10 were N content and fungal to bacteria ratio and then number of trees, beta-xylosidases and soil temperature (Figure x).

![Figure 8. Principal component regression between edaphic and ecological variables and temperature sensitivity (Q10) analysed by Partial least squares (PLS) regression (upper pannel). Variance decomposition of main variables explaining Q10 (lower pannel)](../Figures/Q10_PLS_biplot.png){width=60%}

@@

## Role of temperature and moisture in decomposition
Our model showed that soil temperature and moisture were able to explained more than 50% of variance in soil respiration. Because phenologial patterns in deciduous species can affect sensitivity of soil respiration to temperature (Curiel-Yuste et al 2004), in this model we have also included a seasonality scaling. The seasonal scaling although apparently differentiated slightly between sites, it always presented very low amplitude posteriors, weighting on average only a 5% of the total measured respiration. It was more important in France (between 4.5% and 12.2%) and much less in Spain and Romania (between 2.0% and 3.2%). This is probably related to higher photosynthetic activity variation of Quercus pubescens growing in a Mediterranean climate in France. This would indicate higher changes in root respiration and proliferation that would in turn affect microbial communities. Because in the three study sites there were deciduous Quercus species, another explanation is that soil respiration could have been also affected by the deciduousness of the understorey vegetation (Curiel-Yuste 2004).

## Forest management effect on activation energy and sensitivity to temperature
Our results confirmed the hypothesis 2, since addition of slash decreased the activation energy probably related to a reduction in chemical recalcitrance. This supports the carbon quality-temperature hypothesis stating that the biochemically recalcitrant C pool is more sensitive to changes in temperature than the labile C pool (Bossata and Agren 1999). We also observed a positive correlation between Q10 and activation energy agreeing with Davidson & Janssens, 2006 and Sierra, 2012 that the higher the activation energy the higher is the relative temperature sensitivity.

In line with our hypothesis x, we have observed a general negative trend of activation energy with time in all management disturbance treatments and all sites. This result also gives support to the C quality temperature hypothesis since fresh inputs from the new vegetation low down the temperature sensitivity of C mineralization as these inputs are more labile to decomposed (Lefevre et al 2014). Still, the growth of new vegetation, within three years from disturbance, did not compensate for the decreases of activation energies in the plots where removal of slash had an effect. Additionally, we cannot rule out the fact that other mechanisms could have contributed to low activation energies with time such as differences between years in primary production (i.e., contribution of autotrophic respiration) or changes in substrate and enzyme diffusivity driven by differences in soil moisture between years (Davinson and Jansen 2022). In line with this, for example, in the control plots from Spain the activation energy decreased from year zero (2021) to year two (2023) and then it increased a bit while in disturbed plots was always descendent. Since changes in activation energy due to the growth of new vegetation is not expected in undisturbed control plots and given that 2022 and 2023 were extremely dry in Spain, probably drought exerted a constrained access of enzymes to substrates (Davinson and Jansen 2022). This suggest that drought effect may have similar impacts on temperature sensitivity as chemical recalcitrance (Levegre et al 2014).

## Role of edaphic, ecological and microbiological variables in temperature sensitivity
The Q10 values obtained in our study were comparable to the Q10 values measured in the field by other studies (ref). Xu et al 2015 observed the largest relative frequency of Q10 values across forest in China was within the range of 2.0–3.0. Fierer et al. (2006) observed Q10 values ranging from 2.2 to 4.6 (average of 3.0) at continental-scale in an incubation experiment.

Hydrologic leaching of dissolved and particulate carbon compounds can also be important outputs of C from the system (Davinson and Jansen 2006), that could explained the low carbon content in Romania (sandy soil with low retention capacity). Longer time of frozen soils could reduce the diffusion of substrates and extracellular enzymes within the soil (Davinson and Jansen 2006), given higher values of Q10 not related to recalcitrance.

Our analyses revealed the three variables that explained most of the variation of Q10 were related to i) the quality of substrate (i.e., nitrogen content as limiting nutrient in forest), ii) the microorganisms, in particular, fungal to bateria ratio and iii) the primary productivity (that is, the number of trees).

@@

# Appendix

![Figure S1. Temperature scaling function](../Figures/Appendix/Temp_function.png){width=50%}

![Figure S2. Moisture scaling function](../Figures/Appendix/Moist_function.png){width=50%}

![Figure S3. Seasonality scaling function](../Figures/Appendix/Season_function.png){width=50%}

![Figure S4. Boxplots of temperature sensitivity (Q10) of soil respiration per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/posteriors_boxplots_q10.png){width=60%}

![Figure S5. Posteriors of the scaling parameter, varying as a function of single plots](../Figures/Appendix/posteriors_boxplots_A.png){width=60%}

![Figure S6. Posteriors of the parameters of the temperature function, plotted as density distributions](../Figures/Appendix/posteriors_temp.png)

![Figure S7. Posteriors of the parameters of the moisture function, plotted as density distributions](../Figures/Appendix/posteriors_seasonality.png)

![Figure S8. Posteriors of the parameters of the seasonality function, plotted as density distributions](../Figures/Appendix/posteriors_moisture.png)

![Figure S9. Ratio (expressed as %) between the mean amplitude parameter and the mean measured respiration by treatment](../Figures/Amplitude_ratio.png)
